---
title: "data_manipulation"
author: "Pei Liu"
date: "2022-09-22"
output: html_document
---



```{r}
library(tidyverse)

options(tibble.print_min = 3)

litters_data = read_csv("data_import_examples/FAS_litters.csv",
  col_types = "ccddiiii")
litters_data = janitor::clean_names(litters_data)

pups_data = read_csv("data_import_examples/FAS_pups.csv",
  col_types = "ciiiii")
pups_data = janitor::clean_names(pups_data)

```


select
For a given analysis, you may only need a subset of the columns in a data table; extracting only what you need can helpfully de-clutter, especially when you have large datasets. Select columns using select.

You can specify the columns you want to keep by naming all of them:
```{r}
select(litters_data, group, litter_number, gd0_weight, pups_born_alive)
```
You can specify the specify a range of columns to keep:
```{r}
select(litters_data, group:gd_of_birth)
```
You can also specify columns you’d like to remove:
```{r}
select(litters_data, -pups_survive)
```
You can rename variables as part of this process (but noted that we don't actually change the column name stored in litters data):
```{r}
select(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```
If all you want to do is rename something, you can use rename instead of select. This will rename the variables you care about, and keep everything else:
```{r}
rename(litters_data, GROUP = group, LiTtEr_NuMbEr = litter_number)
```

There are some handy helper functions for select; read about all of them using ?select_helpers. I use starts_with(), ends_with(), and contains() often, especially when there variables are named with suffixes or other standard patterns:
```{r}
select(litters_data, starts_with("gd"))
```
I also frequently use is everything(), which is handy for reorganizing columns without discarding anything:
```{r}
select(litters_data, litter_number, pups_survive, everything())
```
relocate does a similar thing (and is sort of like rename in that it’s handy but not critical):
```{r}
relocate(litters_data, litter_number, pups_survive)
```
In larger datasets,

Lastly, like other functions in dplyr, select will export a dataframe even if you only select one column. Mostly this is fine, but sometimes you want the vector stored in the column. To pull a single variable, use pull.


