---
title: "tidydata_and_relationaldatabase"
author: "Pei Liu"
date: "2022-09-27"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

load some relevant packages, and limit the number of lines printed in a tibble
```{r}
library(tidyverse)

options(tibble.print_min = 5)

```

Load the data
```{r}
pulse_df = 
  haven::read_sas("data_import_examples/public_pulse_data.sas7bdat") %>%
  janitor::clean_names()

pulse_df
```

pivot longer the data
```{r}
pulse_tidy_data = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    values_to = "bdi")

pulse_tidy_data
```

pivot longer the data and Remove the prefix of "visit"
```{r}
pulse_tidy_data = 
  pivot_longer(
    pulse_df, 
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi")

pulse_tidy_data
```
Relocate: to change the column order,
mutate
import, tidy, and transform the PULSE dataset into a usable format:
```{r}
pulse_df = 
  haven::read_sas("data_import_examples/public_pulse_data.sas7bdat") %>%
  janitor::clean_names() %>%
  pivot_longer(
    bdi_score_bl:bdi_score_12m,
    names_to = "visit", 
    names_prefix = "bdi_score_",
    values_to = "bdi") %>%
  relocate(visit) %>%
  mutate(
    visit = replace(visit, visit == "bl", "00m"),
    visit = factor(visit)) %>%
  arrange(id, visit)

print(pulse_df, n = 12)
```


Learning Assessment: In the litters data, the variables gd0_weight and gd18_weight give the weight of the mother mouse on gestational days 0 and 18. Write a data cleaning chain that retains only litter_number and these columns; produces new variables gd and weight; and makes gd a numeric variable taking values 0 and 18 (for the last part, you might want to use recode …). Is this version “tidy”?


```{r}
litter_df = 
  read_csv("data_import_examples/FAS_litters.csv")
janitor::clean_names(litter_df)
litter_df
```



```{r}
litters_wide = 
  read_csv("data_import_examples/FAS_litters.csv") %>%
  janitor::clean_names() %>%
  select(litter_number, gd0_weight, gd18_weight) %>% 
  pivot_longer(
    gd0_weight:gd18_weight,
    names_to = "gd", 
    values_to = "weight") %>% 
  mutate(gd = recode(gd, "gd0_weight" = 0, "gd18_weight" = 18))
litters_wide
```


creates a tidy dataset that could result from an analysis. This is the correct format for additional analysis or visualization, but doesn’t facilitate quick comparisons for human readers:
```{r}
analysis_result = tibble(
  group = c("treatment", "treatment", "placebo", "placebo"),
  time = c("pre", "post", "pre", "post"),
  mean = c(4, 8, 3.5, 4)
)

analysis_result
```

```{r}
pivot_wider(
  analysis_result, 
  names_from = "time", 
  values_from = "mean")
```

To produce the desired tidy dataset, we first need to read each table and do some cleaning.
```{r}
fellowship_ring = 
  readxl::read_excel("data_import_examples/LotR_Words.xlsx", range = "B3:D6") %>%
  mutate(movie = "fellowship_ring")
fellowship_ring


two_towers = 
  readxl::read_excel("data_import_examples/LotR_Words.xlsx", range = "F3:H6") %>%
  mutate(movie = "two_towers")
two_towers

return_king = 
  readxl::read_excel("data_import_examples/LotR_Words.xlsx", range = "J3:L6") %>%
  mutate(movie = "return_king")
return_king
```


```{r}
lotr_tidy = 
  bind_rows(fellowship_ring, two_towers, return_king) %>%
  janitor::clean_names() %>%
  pivot_longer(
    female:male,
    names_to = "gender", 
    values_to = "words") %>%
  mutate(race = str_to_lower(race)) %>% 
  select(movie, everything()) 

lotr_tidy
```
Joining data
```{r}
pup_data = 
  read_csv("data_import_examples/FAS_pups.csv") %>%
  janitor::clean_names() %>%
  mutate(
    sex = recode(sex, `1` = "male", `2` = "female"),
    sex = factor(sex)) 
pup_data
```

```{r}
litter_data = 
  read_csv("data_import_examples/FAS_litters.csv") %>%
  janitor::clean_names() %>%
  separate(group, into = c("dose", "day_of_tx"), sep = 3) %>%
  relocate(litter_number) %>%
  mutate(
    wt_gain = gd18_weight - gd0_weight,
    dose = str_to_lower(dose))
litter_data
```

left_join includes all the columns on the left
```{r}
fas_data = 
  left_join(pup_data, litter_data, by = "litter_number")

fas_data
```
Learning Assessment: The datasets in this zip file contain de-identified responses to surveys included in past years of this course. Both contain a unique student identifier; the first has responses to a question about operating systems, and the second has responses to questions about degree program and git experience. Write a code chunk that imports and cleans both datasets, and then joins them.

Solution:
```{r}
os_df = read_csv("survey_results/surv_os.csv")
os_df
git_exp_df = read_csv("survey_results/surv_program_git.csv")
git_exp_df
```

